---
title: "NY Shootings and Disadvantaged Areas"
output:
  html_document: default
  pdf_document: default
date: "2025-02-17"
---

# Neighborhood Disadvantage and Gun Violence in NYC: A 2023 Analysis


## Introduction
This study examines the relationship between neighborhood disadvantage scores and gun violence
in New York City in 2023. With ongoing discussions about public safety and crime rates, this
research provides a data-driven approach to understanding the link between socioeconomic
conditions and shootings at the neighborhood level.
Using shooting data from the New York Police Department (NYPD) and disadvantaged
neighborhood classifications from the New York State Energy Research and Development Authority
(NYSERDA), this study seeks to determine whether neighborhoods with higher disadvantage scores
experience more gun violence. The findings reveal a strong correlation between these two factors,
indicating that disadvantaged communities face disproportionately higher rates of shootings per
100,000 people.


```{r,include=FALSE}
rm(list = ls())
gc()

options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages(c("sf", "ggplot2", "dplyr"))


# Set working directory
directory <- "C:/Users/mikem/iCloudDrive/NYC Shootings/NYC Shooting Project Data Sets"
setwd(directory)

# Set seed for reproducibility
set.seed(123)

# Load required libraries
if (!require("here")) install.packages("here", dependencies = TRUE)

# Load datasets
shootings <- read.csv("NYPD_Shooting_Data.csv", stringsAsFactors = FALSE)
neighborhoods <- read.csv("2023_NY_Disadvantaged Neighborhoods.csv", stringsAsFactors = FALSE)
census <- read.csv("NYC Census Tracts with Shootings.csv", stringsAsFactors = FALSE)
cen_population <- read.csv("Cleaned_Census_Tract_Populations.csv", stringsAsFactors = FALSE)
per100 <- read.csv("Shootings_Table_With_Population_Disadvantage.csv", stringsAsFactors = FALSE)
install.packages(c("sf", "ggplot2", "dplyr"))
```
## Data Sources

This analysis relies on two primary datasets:

1. NYPD Shooting Incident Data (2023):
  • Obtained from the New York Police Department (NYPD), this dataset provides
detailed information on all recorded shootings in NYC from 2006 to 2023.
  • Includes variables such as incident date/time, location (latitude/longitude), victim
demographics (age, race, sex), and whether the shooting was fatal or non-fatal.
  • This dataset enables the calculation of shooting rates per 100,000 residents at the
neighborhood level by matching incidents to NYC census tracts.

2. New York State Disadvantaged Communities Data (2023): Defines disadvantaged
neighborhoods based on environmental burdens, climate risks, and socioeconomic
vulnerabilities.
  • Published by the New York State Energy Research and Development Authority
(NYSERDA), this dataset identifies disadvantaged communities based on a
combination of socioeconomic, environmental, and health-related factors.
  • Neighborhoods are assigned a Disadvantaged Score ranging from 0 (least
disadvantaged) to 1 (most disadvantaged), determined by indicators such as
poverty rates, unemployment, median income, racial composition, housing burden,
and environmental risks.
  • This dataset allows for a quantitative measure of neighborhood-level disadvantage,
which is used to assess its relationship with gun violence.


By integrating these two datasets, this study evaluates whether higher levels of neighborhood
disadvantage correlate with increased rates of gun violence.


# Methodology

1. Geo-Spatial Analysis: Mapping Shootings to Census Tracts
  • Since both datasets use different geographic units (point-level data for shootings vs.
census tract-level data for disadvantage), shootings were spatially matched to NYC
census tracts using GIS tools and R’s sf package.
  • The total number of shootings within each census tract was calculated, allowing for
neighborhood-level comparison.

# This map shows NYC most disadvantaged neighborhoods in 2023 along with their demographic makeup
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(sf)
library(leaflet)
library(dplyr)


# Convert WKT geometry to sf object
neighborhoods_sf <- st_as_sf(neighborhoods, wkt = "the_geom", crs = 4326)

# Ensure geometries are valid
neighborhoods_sf <- st_make_valid(neighborhoods_sf)

# Filter for NYC bounding box (approximate boundaries)
nyc_sf <- neighborhoods_sf %>%
  st_filter(st_as_sfc(st_bbox(c(xmin = -74.2591, xmax = -73.7004, ymin = 40.4774, ymax = 40.9176), crs = 4326)))

# Filter only areas that are "Designated as DAC"
nyc_disadvantaged_sf <- nyc_sf %>%
  filter(DAC_Designation == "Designated as DAC")

# Check if there are valid rows left
if (nrow(nyc_disadvantaged_sf) == 0) {
  stop("No areas found with DAC_Designation == 'Designated as DAC'. Verify column values.")
}

# Create an interactive leaflet map highlighting DAC-designated areas
leaflet(nyc_disadvantaged_sf) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(
    fillColor = "lightblue",
    color = "blue",
    weight = 1,
    opacity = 0.7,
    fillOpacity = 0.6,
    highlightOptions = highlightOptions(
      color = "yellow", weight = 2, bringToFront = TRUE
    ),
    label = lapply(
      paste0(
        "<div style='font-size:14px; background-color:white; padding:5px; border-radius:5px; border:1px solid #ccc;'>",
        "<b>GEOID:</b> ", nyc_disadvantaged_sf$GEOID, "<br>",
        "<b>County:</b> ", nyc_disadvantaged_sf$County, "<br>",
        "<b>Percentile Rank Combined NYC:</b> ", nyc_disadvantaged_sf$Percentile_Rank_Combined_NYC, "<br>",
        "<b>Black/African American Percent:</b> ", nyc_disadvantaged_sf$Black_African_American_Percent, "%<br>",
        "<b>Latino Percent:</b> ", nyc_disadvantaged_sf$Latino_Percent, "%",
        "</div>"
      ),
      htmltools::HTML
    ),
    labelOptions = labelOptions(
      direction = "auto",
      textsize = "12px",
      style = list(
        "color" = "black",
        "background-color" = "white",
        "border" = "1px solid gray",
        "padding" = "6px",
        "border-radius" = "6px"
      )
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors = "lightblue",
    labels = "Most Disadvantaged Neighborhoods",
    opacity = 1.0
  ) %>%
  setView(lng = -74.0060, lat = 40.7128, zoom = 11)

```

# This map shows all NYC shootings since 2006 along with statistical markers for each shooting
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(leaflet)
library(dplyr)
library(lubridate)  # For date handling

# Inspect column names to confirm latitude/longitude fields
print(names(shootings))

# Create a data frame for coordinates with additional statistical markers
location.data <- shootings %>%
  filter(!is.na(Latitude) & !is.na(Longitude)) %>%  # Remove rows with missing coordinates
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    OCCUR_DATE = mdy(OCCUR_DATE),  # Convert date format to readable format
    STATISTICAL_MURDER_FLAG = ifelse(STATISTICAL_MURDER_FLAG, "Fatal", "Non-Fatal") # Convert to readable format
  ) %>%     
  select(Longitude, Latitude, OCCUR_DATE, OCCUR_TIME, PRECINCT, 
         STATISTICAL_MURDER_FLAG, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)

# NYC Map with shooting data plotted
leaflet(data = location.data) %>%
  addProviderTiles("CartoDB.Positron") %>%  # Cleaner map style
  addCircleMarkers(
    ~Longitude, ~Latitude,
    radius = 2,  # Adjust marker size for visibility
    color = "red",
    fillColor = "red",
    fillOpacity = 0.2,
    popup = ~paste0(
      "<b>Date:</b> ", OCCUR_DATE, "<br>",
      "<b>Time:</b> ", OCCUR_TIME, "<br>",
      "<b>Precinct:</b> ", PRECINCT, "<br>",
      "<b>Outcome:</b> ", STATISTICAL_MURDER_FLAG, "<br>",
      "<b>Victim Age Group:</b> ", VIC_AGE_GROUP, "<br>",
      "<b>Victim Sex:</b> ", VIC_SEX, "<br>",
      "<b>Victim Race:</b> ", VIC_RACE)
  ) %>%
  setView(lng = -74.0060, lat = 40.7128, zoom = 12)


```

# Heatmap of the NYC Shooting Data since 2006 Showing areas of hightened gun violence
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(leaflet)
library(leaflet.extras)  # Enables heatmap functionality
library(dplyr)
library(lubridate)  # For date handling

# Process data: filter and format relevant columns
location.data <- shootings %>%
  filter(!is.na(Latitude) & !is.na(Longitude)) %>%  # Remove rows with missing coordinates
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    OCCUR_DATE = mdy(OCCUR_DATE)  # Convert date format
  ) %>%
  select(Longitude, Latitude, OCCUR_DATE)

# NYC Map with a red-intense heatmap layer
leaflet(data = location.data) %>%
  addProviderTiles("CartoDB.Positron") %>%  # Cleaner map style
  addHeatmap(
    lng = ~Longitude, lat = ~Latitude,
    intensity = 7,  # Each point contributes equally to the heatmap
    blur = 8,  # Decrease blur for sharper heat spots
    max = 15,  # Increase max intensity for a stronger color contrast
    radius = 25,  # Larger radius for more concentrated heat spots
    gradient = c("transparent", "#ffd700", "#ffa500", "#990000")  # Gradient shades of red
  ) %>%
  setView(lng = -74.0060, lat = 40.7128, zoom = 12)

```

# This map shows all NYC shootings in 2023 along with statistical markers for each shooting
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(leaflet)
library(dplyr)
library(lubridate)  # For easier date handling

# Inspect column names to confirm date and latitude/longitude fields
print(names(shootings))

# Convert OCCUR_DATE to Date format and filter for 2023
location.data <- shootings %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%  # Convert from MM/DD/YYYY to Date format
  filter(year(OCCUR_DATE) == 2023 & !is.na(Latitude) & !is.na(Longitude)) %>%  
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    STATISTICAL_MURDER_FLAG = ifelse(STATISTICAL_MURDER_FLAG, "Fatal", "Non-Fatal") # Convert to readable format
  ) %>%     
  select(Longitude, Latitude, OCCUR_DATE, OCCUR_TIME, PRECINCT, STATISTICAL_MURDER_FLAG, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)

# NYC Map with shooting data plotted
leaflet(data = location.data) %>%
  addProviderTiles("CartoDB.Positron") %>%  # Cleaner map style
  addCircleMarkers(
    ~Longitude, ~Latitude,
    radius = 4,  # Adjust marker size for better visibility
    color = "red",
    fillColor = "red",
    fillOpacity = 0.5,
    popup = ~paste(
      "<b>Date:</b>", OCCUR_DATE, "<br>",
      "<b>Time:</b>", OCCUR_TIME, "<br>",
      "<b>Precinct:</b>", PRECINCT, "<br>",
      "<b>Outcome:</b>", STATISTICAL_MURDER_FLAG, "<br>",
      "<b>Victim Age Group:</b>", VIC_AGE_GROUP, "<br>",
      "<b>Victim Sex:</b>", VIC_SEX, "<br>",
      "<b>Victim Race:</b>", VIC_RACE, "<br>"
    )
  ) %>%
  setView(lng = -74.0060, lat = 40.7128, zoom = 12)


```

# This map displays both shootings and most Disadvantaged Neighborhoods in 2023
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(sf)
library(leaflet)
library(dplyr)
library(lubridate)  # For date handling

# --- Process Shooting Data ---
# Convert OCCUR_DATE to Date format and filter for 2023 shootings
shooting_data <- shootings %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%  # Convert from MM/DD/YYYY to Date format
  filter(year(OCCUR_DATE) == 2023 & !is.na(Latitude) & !is.na(Longitude)) %>%  
  mutate(
    Latitude = as.numeric(Latitude),
    Longitude = as.numeric(Longitude),
    STATISTICAL_MURDER_FLAG = ifelse(STATISTICAL_MURDER_FLAG, "Fatal", "Non-Fatal") # Convert to readable format
  ) %>%     
  select(Longitude, Latitude, OCCUR_DATE, OCCUR_TIME, PRECINCT, 
         STATISTICAL_MURDER_FLAG, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)

# --- Process DAC Data ---
# Convert WKT geometry to sf object
neighborhoods_sf <- st_as_sf(neighborhoods, wkt = "the_geom", crs = 4326)

# Ensure geometries are valid
neighborhoods_sf <- st_make_valid(neighborhoods_sf)

# Filter for NYC bounding box
nyc_sf <- neighborhoods_sf %>%
  st_filter(st_as_sfc(st_bbox(c(xmin = -74.2591, xmax = -73.7004, ymin = 40.4774, ymax = 40.9176), crs = 4326)))

# Filter only areas that are "Designated as DAC"
nyc_disadvantaged_sf <- nyc_sf %>%
  filter(DAC_Designation == "Designated as DAC")

# Check if there are valid rows left
if (nrow(nyc_disadvantaged_sf) == 0) {
  stop("No areas found with DAC_Designation == 'Designated as DAC'. Verify column values.")
}

# --- Combine Maps ---
leaflet() %>%
  # Add base map
  addProviderTiles("CartoDB.Positron") %>%

  # Add DAC-designated neighborhoods as polygons
  addPolygons(data = nyc_disadvantaged_sf,
    fillColor = "lightblue",
    color = "blue",
    weight = 1,
    opacity = 0.7,
    fillOpacity = 0.6,
    group = "Disadvantaged Neighborhoods",
    highlightOptions = highlightOptions(
      color = "yellow", weight = 2, bringToFront = TRUE
    ),
    label = ~paste("DAC Status:", DAC_Designation)
  ) %>%

  # Add shooting incidents as red circle markers with detailed popups
  addCircleMarkers(data = shooting_data,
    ~Longitude, ~Latitude,
    radius = 2,  # Adjust marker size for visibility
    color = "red",
    fillColor = "red",
    fillOpacity = 0.1,
    group = "Shooting Incidents",
    popup = ~paste0(
      "<b>Date:</b> ", OCCUR_DATE, "<br>",
      "<b>Time:</b> ", OCCUR_TIME, "<br>",
      "<b>Precinct:</b> ", PRECINCT, "<br>",
      "<b>Outcome:</b> ", STATISTICAL_MURDER_FLAG, "<br>",
      "<b>Victim Age Group:</b> ", VIC_AGE_GROUP, "<br>",
      "<b>Victim Sex:</b> ", VIC_SEX, "<br>",  # Missing comma added
      "<b>Victim Race:</b> ", VIC_RACE  # This was missing, added back
    )
  ) %>%

  # Add layer control to toggle between datasets
  addLayersControl(
    overlayGroups = c("Disadvantaged Neighborhoods", "Shooting Incidents"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%

  # Add legends for clarity
  addLegend(
    position = "bottomleft",
    colors = c("lightblue", "red"),
    labels = c("Disadvantaged Neighborhoods", "Shooting Incidents"),
    opacity = 1.0
  ) %>%

  # Center the map on NYC
  setView(lng = -74.0060, lat = 40.7128, zoom = 12)

```



# This Table displays the top 15 Cenus Tracts with the highest disadvantage score in NYC 2023
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(knitr)
library(kableExtra)

# Define NYC boroughs
nyc_boroughs <- c("Bronx", "Manhattan", "Queens", "Brooklyn", "Staten Island")

# Filter for NYC neighborhoods
nyc_data <- neighborhoods %>%
  filter(County %in% nyc_boroughs)

# Select relevant columns
relevant_columns <- c("GEOID", "County", "Percentile_Rank_Combined_NYC")

# Get the top 15 census tracts with the highest disadvantaged scores
top_15_disadvantaged <- nyc_data %>%
  select(all_of(relevant_columns)) %>%
  arrange(desc(Percentile_Rank_Combined_NYC)) %>%
  head(15)

# Display the table with formatting
top_15_disadvantaged %>%
  mutate(GEOID = paste0("**", GEOID, "**")) %>%  # Make GEOID bold
  kable("html", caption = "<span style='font-weight: bold; color: black;'>Top 15 Most Disadvantaged Census Tracts in NYC</span>") %>%
  kable_styling("striped", full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, color = "red", bold = TRUE)  # Highlight disadvantaged scores in red

```


# ==========================================
## Understanding `Percentile_Rank_Combined_NYC`

### **What Does This Score Represent?**
The `Percentile_Rank_Combined_NYC` variable represents the **relative disadvantaged ranking of a census tract within New York City (NYC)**. 

- **Range: 0 to 1** (or 0% to 100%)
- **Higher values** (closer to **1.00**) indicate that the neighborhood is **more disadvantaged** relative to other NYC census tracts.
- **Lower values** (closer to **0.00**) indicate that the neighborhood is **less disadvantaged** compared to other NYC areas.

---

### **How Is It Calculated?**
NYC assigns this percentile rank based on a **composite index** that combines **two major factors**:

1. **Environmental Burden Score**:
   - Measures **pollution exposure** (e.g., air quality, hazardous waste sites, toxic releases).
   - Includes **climate risks** (e.g., flood vulnerability, extreme heat).
   
2. **Population Vulnerability Score**:
   - Includes **socioeconomic factors** (e.g., poverty rate, education levels, housing burden).
   - Considers **health risks** (e.g., asthma rates, access to healthcare).

Each census tract is **ranked based on these combined scores**, and the **percentile rank is assigned accordingly**.

---

### **Interpretation of the Score**
The table below explains how to interpret the percentile rank:

| `Percentile_Rank_Combined_NYC` | **Interpretation** |
|-----------------------------|----------------------|
| **1.00 (100%)** | **Most Disadvantaged** Neighborhood in NYC |
| **0.90 (90%)** | Among **Top 10%** Most Disadvantaged Areas |
| **0.50 (50%)** | **Median** Level of Disadvantagedness |
| **0.10 (10%)** | Among **Least Disadvantaged** Areas |
| **0.00 (0%)**  | **Least Disadvantaged** Neighborhood in NYC |

For example:
- A neighborhood with a **percentile rank of 0.98** means that it is **more disadvantaged than 98% of all census tracts in NYC**.
- A neighborhood with a **percentile rank of 0.25** means that it is **less disadvantaged than 75% of census tracts**.

---

### **Why is This Score Important?**
- Helps **identify neighborhoods that need the most resources** (e.g., funding, clean energy initiatives, public health interventions).
- Used to **prioritize disadvantaged communities** for **government programs and policy planning**.
- Supports **environmental justice** efforts by addressing communities with **high pollution and socioeconomic struggles**.

---

### **Example Interpretation**

Consider two neighborhoods:
A **Bronx census tract** with:
- **GEOID: 36005009300**
- **Percentile_Rank_Combined_NYC: 1.00 (100%)**  
👉 This means **it is the most disadvantaged neighborhood in all of NYC**.

A **Manhattan census tract** with:
- **GEOID: 36061000100**
- **Percentile_Rank_Combined_NYC: 0.15 (15%)**  
👉 This means **it is among the least disadvantaged neighborhoods in NYC**.

---

### **Key Takeaways**
✔ The **higher the percentile rank**, the **more disadvantaged** the neighborhood.  
✔ The score is **relative to NYC**, so it **compares neighborhoods only within the city**.  
✔ Used by **policymakers and researchers** to **target resources where they are needed most**.





# This Table displays the top 15 Most Disadvantaged Census Tracts in NYC (2023) with Shooting Counts
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(sf)
library(lubridate)
library(knitr)
install.packages("kableExtra")
library(kableExtra)

# Define NYC boroughs
nyc_boroughs <- c("Bronx", "Manhattan", "Queens", "Brooklyn", "Staten Island")

# Ensure OCCUR_DATE is in Date format and filter for 2023 shootings
shootings_2023 <- shootings %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%  # Convert to Date format
  filter(year(OCCUR_DATE) == 2023 & !is.na(Latitude) & !is.na(Longitude))

# Convert shootings data to spatial points using sf package
shootings_sf <- st_as_sf(shootings_2023, coords = c("Longitude", "Latitude"), crs = 4326)

# Convert neighborhoods dataset to sf object
neighborhoods_sf <- st_as_sf(neighborhoods, wkt = "the_geom", crs = 4326) %>%
  filter(County %in% nyc_boroughs)

# Perform spatial join to assign each shooting to a census tract
shootings_with_tracts <- st_join(shootings_sf, neighborhoods_sf, left = FALSE) %>%
  st_drop_geometry()  # Remove geometry column

# Group shootings by census tract
shootings_count <- shootings_with_tracts %>%
  group_by(GEOID) %>%
  summarise(Shooting_Count = n())

# Filter for NYC neighborhoods
nyc_data <- neighborhoods %>%
  filter(County %in% nyc_boroughs)

# Select relevant columns and join with shooting counts
relevant_columns <- c("GEOID", "County", "Percentile_Rank_Combined_NYC")

top_15_disadvantaged <- nyc_data %>%
  select(all_of(relevant_columns)) %>%
  left_join(shootings_count, by = "GEOID") %>%
  mutate(Shooting_Count = ifelse(is.na(Shooting_Count), 0, Shooting_Count)) %>%
  arrange(desc(Percentile_Rank_Combined_NYC)) %>%
  head(15)

# Display the table with formatting
top_15_disadvantaged %>%
  mutate(GEOID = paste0("**", GEOID, "**")) %>%  # Make GEOID bold
  kable("html", caption = "<span style='font-weight: bold; color: black;'>Top 15 Most Disadvantaged Census Tracts in NYC (with Shooting Counts)</span>") %>%
  kable_styling("striped", full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, color = "red", bold = TRUE) %>%  # Highlight disadvantaged scores in red
  column_spec(4, color = "blue", bold = TRUE)  # Highlight shooting counts in blue

```

# This Table Displays All NYC Census Tracts in NYC during 2023 that had atleast 1 intcident of gun violence
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(sf)
library(lubridate)
library(knitr)
library(kableExtra)

# Ensure OCCUR_DATE is in Date format and filter for 2023 shootings
shootings_2023 <- shootings %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%  # Convert to Date format
  filter(year(OCCUR_DATE) == 2023 & !is.na(Latitude) & !is.na(Longitude))

# Convert shootings data to spatial points using sf package
shootings_sf <- st_as_sf(shootings_2023, coords = c("Longitude", "Latitude"), crs = 4326)

# Convert neighborhoods dataset to sf object
neighborhoods_sf <- st_as_sf(neighborhoods, wkt = "the_geom", crs = 4326)

# Perform spatial join to assign each shooting to a census tract
shootings_with_tracts <- st_join(shootings_sf, neighborhoods_sf, left = FALSE) %>%
  st_drop_geometry()  # Remove geometry column

# Group shootings by census tract
shootings_by_tract <- shootings_with_tracts %>%
  group_by(GEOID, County) %>%
  summarise(
    Shooting_Count = n(),  # Count number of shootings per tract
    Disadvantaged_Score = first(Percentile_Rank_Combined_NYC)  # Get disadvantaged score
  ) %>%
  arrange(desc(Shooting_Count))  # Sort by most shootings

# Select only the top 15 census tracts with the most shootings
top_15_shootings <- shootings_by_tract %>%
  top_n(15, Shooting_Count)

# Display the table with the top 15 census tracts
top_15_shootings %>%
  knitr::kable("html", caption = "<span style='font-weight: bold; color: black;'>Census Tracts with at least 1 Instance of Gun Violence (2023)</span>") %>%
  kableExtra::kable_styling("striped", full_width = F, position = "center") %>%
  kableExtra::column_spec(3, color = "red", bold = TRUE)

```

# This Table displays Census Tracts with 0 Shootings in NYC during 2023
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(dplyr)
library(knitr)
library(kableExtra)

# Define NYC boroughs
nyc_boroughs <- c("Bronx", "Manhattan", "Queens", "Brooklyn", "Staten Island")

# Filter for NYC neighborhoods
nyc_data <- neighborhoods %>%
  filter(County %in% nyc_boroughs)

# Select relevant columns
relevant_columns <- c("GEOID", "County", "Percentile_Rank_Combined_NYC")

# Get the top 15 census tracts with the lowest disadvantaged scores
lowest_15_disadvantaged <- nyc_data %>%
  select(all_of(relevant_columns)) %>%
  arrange(Percentile_Rank_Combined_NYC) %>%  # Arrange in ascending order for lowest scores
  head(15)

# Display the table with formatting
lowest_15_disadvantaged %>%
  mutate(GEOID = paste0("**", GEOID, "**")) %>%  # Make GEOID bold
  kable("html", caption = "<span style='font-weight: bold; color: black;'>Top 15 Least Disadvantaged Census Tracts in NYC</span>") %>%
  kable_styling("striped", full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(3, color = "green", bold = TRUE)  # Highlight lowest scores in green
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
# Load necessary libraries
library(dplyr)
library(sf)
library(lubridate)
library(tidyr)
library(knitr)
library(kableExtra)

# Ensure OCCUR_DATE is in Date format and filter for 2023 shootings
shootings_2023 <- shootings %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  filter(year(OCCUR_DATE) == 2023 & !is.na(Latitude) & !is.na(Longitude))

# Convert shootings data to spatial points using sf package
shootings_sf <- st_as_sf(shootings_2023, coords = c("Longitude", "Latitude"), crs = 4326)

# Convert neighborhoods dataset to sf object
neighborhoods_sf <- st_as_sf(neighborhoods, wkt = "the_geom", crs = 4326)

# Filter neighborhoods to only include those in NYC (using corrected county names)
nyc_neighborhoods_sf <- neighborhoods_sf %>%
  filter(County %in% c("Queens", "Brooklyn", "Bronx", "Richmond", "New York"))  # Richmond = Staten Island, New York = Manhattan

# Perform spatial join to assign each shooting to a census tract
shootings_with_tracts <- st_join(shootings_sf, nyc_neighborhoods_sf, left = FALSE) %>%
  st_drop_geometry()  # Remove geometry column

# Group shootings by census tract
shootings_by_tract <- shootings_with_tracts %>%
  group_by(GEOID) %>%
  summarise(
    Shooting_Count = n()  # Count number of shootings per tract
  )

# Get all census tracts, county data, and disadvantaged scores from neighborhoods dataset
all_tracts <- nyc_neighborhoods_sf %>%
  st_drop_geometry() %>%
  select(GEOID, County, Percentile_Rank_Combined_NYC)

# Identify tracts with zero shootings and include county and disadvantaged score
tracts_zero_shootings <- all_tracts %>%
  left_join(shootings_by_tract, by = "GEOID") %>%
  replace_na(list(Shooting_Count = 0)) %>%
  filter(Shooting_Count == 0)

# Export the table to a CSV file
write.csv(tracts_zero_shootings, "Census_Tracts_Zero_Shootings_2023_NYC.csv", row.names = FALSE)

# Display the table with census tracts that had zero shootings, including county and disadvantaged scores
tracts_zero_shootings %>%
  knitr::kable("html", caption = "<span style='font-weight: bold; color: black;'>Census Tracts with Zero Shootings in NYC (2023)</span>") %>%
  kableExtra::kable_styling("striped", full_width = F, position = "center")

```

```{r, message=FALSE, warning=FALSE, include=FALSE}

library(dplyr)
# Ensure GEOID is treated as a character for merging
neighborhoods$GEOID <- as.character(neighborhoods$GEOID)
census$GEOID <- as.character(census$GEOID)

# Merge population and disadvantage score data with the census data
census_with_details <- left_join(
  census, 
  neighborhoods %>% select(GEOID, Population_Count, Percentile_Rank_Combined_NYC), 
  by = "GEOID"
)

# Check if "Disadvantaged_Score" already exists, and remove duplicate before renaming
if ("Disadvantaged_Score" %in% colnames(census_with_details)) {
  census_with_details <- census_with_details %>% select(-Disadvantaged_Score)
}

# Rename columns for clarity
census_with_details <- census_with_details %>%
  rename(
    Population = Population_Count,
    Disadvantaged_Score = Percentile_Rank_Combined_NYC
  )

# Add a new column: Shootings per 100,000 persons
census_with_details <- census_with_details %>%
  mutate(Shootings_per_100k = ifelse(!is.na(Population) & Population > 0, 
                                     (Shooting_Count / Population) * 100000, 
                                     NA))  # Avoid division by zero

# Create the final table with selected columns
shootings_table <- census_with_details %>%
  select(GEOID, County, Population, Shooting_Count, Disadvantaged_Score, Shootings_per_100k) %>%
  arrange(GEOID)  # Sort by GEOID

# Display the table in R
print(shootings_table)

# Save the table as a CSV file
write.csv(shootings_table, "Shootings_Table_With_Population_Disadvantage.csv", row.names = FALSE)


```

# Scatterplot of Disadvantaged Score vs. Shootings per 100,000
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)

# Remove rows with missing values
per100 <- per100[complete.cases(per100$Disadvantaged_Score, per100$Shootings_per_100k), ]

# Define outlier threshold (95th percentile)
outlier_threshold <- quantile(per100$Shootings_per_100k, 0.95, na.rm = TRUE)

# OPTIONAL: Remove extreme outliers (top 1% of shooting values)
per100_filtered <- per100[per100$Shootings_per_100k <= outlier_threshold, ]

# Create scatter plot with LINEAR regression and adjusted y-axis
ggplot(per100_filtered, aes(x = Disadvantaged_Score, y = Shootings_per_100k)) +
  geom_point(alpha = 0.4, color = "blue", size = 1.5) +  # Reduce opacity to reduce clutter
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2) +  # Linear regression for clear positive trend
  labs(
    title = "Disadvantage Score vs. Shootings per 100k",
    x = "Disadvantaged Score",
    y = "Shootings per 100,000"
  ) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(0, 1)) +  # Ensure full range on x-axis
  coord_cartesian(ylim = c(0, outlier_threshold)) +  # Lower y-axis and remove extreme outliers from view
  theme_minimal()


```
# Scatterplot Explination
Each blue dot represents a geographic area (GEOID).
The X-axis (Disadvantaged Score) → How disadvantaged an area is (higher = more disadvantaged).
The Y-axis (Shootings per 100,000 people) → The number of shootings in that area, adjusted for population size.
The Red Line (Trend Line) → Shows the overall pattern

# Scatterplot Analysis
As the Disadvantaged Score increases, the number of shootings per 100,000 people also increases.
This positive correlation suggests that more disadvantaged areas tend to have higher rates of gun violence.

# Scatterplot Conclusion
Key Takeaway
The scatter plot supports the idea that neighborhoods with higher social disadvantage experience more shootings per 100,000 people.
While there is some variation, the upward trend (red line) indicates that disadvantage is linked to higher gun violence.


# Regression Analysis Table
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries
library(ggplot2)
library(broom)  # For tidy regression results
library(dplyr)  # For table formatting
library(knitr)  # For kable table formatting
library(kableExtra)  # For enhanced table styling

# Read the data
per100 <- read.csv("Shootings_Table_With_Population_Disadvantage.csv", stringsAsFactors = FALSE)

# Remove missing values
per100 <- per100[complete.cases(per100$Disadvantaged_Score, per100$Shootings_per_100k), ]

# Define outlier threshold (95th percentile) to match scatterplot adjustments
outlier_threshold <- quantile(per100$Shootings_per_100k, 0.95, na.rm = TRUE)

# Filter data to remove extreme outliers
per100_filtered <- per100 %>% filter(Shootings_per_100k <= outlier_threshold)

# Run linear regression
model <- lm(Shootings_per_100k ~ Disadvantaged_Score, data = per100_filtered)

# Create a formatted regression table
regression_table <- broom::tidy(model) %>%
  select(term, estimate, std.error, statistic, p.value) %>%
  rename(
    Variable = term,
    Estimate = estimate,
    "Std. Error" = std.error,
    "T-Statistic" = statistic,
    "P-Value" = p.value
  )

# Format the table neatly using kableExtra
regression_table %>%
  kable("html", caption = "Regression Results: Disadvantaged Score vs. Shootings per 100k") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(2:5, width = "3cm") %>%  # Adjust column width
  row_spec(0, bold = TRUE)  # Make header bold

```


# ==========================================
# Interpretation of Regression Results
# ==========================================

#Relationship Between Disadvantaged Score & Shootings per 100k

#Since Disadvantaged Score is a percentage (0 to 1), we interpret the coefficient in percentage point changes.
Each 1% increase in Disadvantaged Score (0.01 increase) is associated with an increase of ~0.30 shootings per 100,000 people.
A 10 percentage point increase (e.g., from 20% to 30%) is linked to ~3 additional shootings per 100,000 people.
A 50 percentage point increase (e.g., from 20% to 70%) is linked to ~15 additional shootings per 100,000 people.
#Conclusion: There is a strong positive correlation between neighborhood disadvantage and gun violence.

# Statistical Significance (T-Statistic & P-Value)

#The T-Statistic (17.32) is very high, indicating that the effect of Disadvantaged Score on Shootings per 100k is strong and unlikely to be random.
The P-Value (0.000) confirms that this relationship is highly statistically significant—meaning the chance that this effect is due to randomness is virtually zero.
#Conclusion: The data strongly supports the idea that higher disadvantage leads to higher rates of shootings per 100,000 people.

#Interpreting the Intercept (-2.30)

#The intercept represents the expected number of shootings per 100,000 when Disadvantaged Score = 0 (i.e., in the least disadvantaged areas).
Since the intercept is negative (-2.30 shootings per 100k), it is not meaningful in a real-world context—shootings cannot be negative.
This happens because the regression line is trying to fit the data, but very few areas have a Disadvantaged Score of exactly 0, making the intercept less relevant.


#Conclusion: We should focus on the slope (30.43 per full unit or ~0.30 per 1%), which provides meaningful insights into how shootings change with increasing disadvantage.

# ==========================================
# 🔹 Final Takeaways
# ==========================================
# ✔ More disadvantaged areas experience significantly higher gun violence.
# ✔ A small increase in disadvantage (e.g., 10%) has a noticeable effect on shootings per 100k.
# ✔ The statistical significance (high T-score & low P-value) confirms this is not due to chance.
# ✔ The intercept is not meaningful, but the slope tells us how much shootings increase with disadvantage.
